---
title: Modular design and software development best practices in R
institute: Odum Institute<br/>University of North Carolina at Chapel Hill
format:
    revealjs:
        theme: [default, unc.scss]
        width: 1920
        height: 1080
        logo: UNC_logo_RGB.png
        slide-number: true
execute: 
  include: true
  echo: true
---

```{r}
#| include: false
#| echo: false

library(tidyverse)

data = read_csv("../data/air_sample.csv")

market_ids = read_csv("../data/L_CITY_MARKET_ID.csv")
data = left_join(data, rename(market_ids, OriginCity="Description"), by=c(OriginCityMarketID="Code"))
data = left_join(data, rename(market_ids, DestCity="Description"), by=c(DestCityMarketID="Code"))

carriers = read_csv("../data/L_CARRIERS.csv")
data = left_join(data, rename(carriers, OperatingCarrierName="Description"), by=c(OpCarrier="Code"))
data = left_join(data, rename(carriers, TicketingCarrierName="Description"), by=c(TkCarrier="Code"))
```

# About me

- Assistant Professor of City and Regional Planning
    - Research focus: transportation modeling and simulation
- Quantitative Methods Consultant at Odum
- Three years as a software developer before graduate school

# Why this matters

- Code is increasingly common in research
- Most academics receive little to no formal training in software development
- This can lead to
    - Inefficient development (takes too much time)
    - Errors in code
    - Non-reproducible analyses

# What we'll be doing today

- Hands-on exercise converting a set of R scripts a modular and tested format
- Two scripts that analyze data on airline tickets from the Bureau of Transportation Statistics
- Script download: [https://github.com/mattwigway/odum-modular-design-r](https://github.com/mattwigway/odum-modular-design-r)
- These slides: [https://projects.indicatrix.org/odum-modular-design-r](https://projects.indicatrix.org/odum-modular-design-r)

# The two scripts

- Both load and clean the air ticket data
- One analyzes the busiest air routes in the country
- One looks at airline "fortress hubs"---which airports have the most concentrated service from a single airline

# What we'll address

- Duplicated code
- Literal numeric values
- Automated tests
- Package version incompatibilities

# What we won't address

- Version control
    - Odum offers separate Git and Github class

# `air_analysis.R`

- Hands-on walkthrough of file

# `airline_analysis.R`

- Hands-on walkthrough of file

# Duplicated code (don't repeat yourself)

- Common software development rule: don't repeat yourself (DRY)
- In `air_analysis.R` we do the same analysis twice, once by airport and once by city
- We can eliminate this duplicated code using a _function_

# Functions

- A function is a reusable piece of code that can be executed many times
- Functions separate the _definition_ of code from the _execution_
- You use functions all the time
    - For instance, `sum()` is a function

# Anatomy of a function

- Functions are R code that is given a name to be referred to later
- Functions have three main components
    - Body
        - The main R code of the function
    - Arguments
        - Values that can be specified each time the function is used
    - Return value
        - The result of the function

# Anatomy of a function


```{r}
calculate_cylinder_volume = function (radius, height) {
    base_area = pi * radius ^ 2
    volume = base_area * height
    return(volume)
}
```

::: {.incremental}
- _radius_ and _height_ are arguments
- Next two lines are body (the actual calculations)
- _volume_ is the return value
    - `return` ends the function, any code after `return` will not be run
:::

# Using functions

- Code above _defines_ the function
- It doesn't run any of the function code
- Function code is run when the function is _called_

. . . 

```{r}
calculate_cylinder_volume(1, 2)
```

# Defining functions in `air_analysis.R`

- The analysis is repeated twice, once by airport pairs and once by city pairs
- We will re-write this analysis into a function, and use the same code twice

# Defining functions in `air_analysis.R`

```{r}
#| output: false
pairs = group_by(data, Origin, Dest) %>% summarize(Passengers=sum(Passengers),
    distance_km=first(Distance) * 1.609)
arrange(pairs, -Passengers)

pairs = mutate(pairs,
    airport1 = if_else(Origin < Dest, Origin, Dest),
    airport2 = if_else(Origin < Dest, Dest, Origin)
    )
pairs = group_by(pairs, airport1, airport2) %>% summarize(
    Passengers=sum(Passengers),
    distance_km=first(distance_km)
)
arrange(pairs, -Passengers)
```

# Step one: make it a function

- Wrap code in `function () {...}` and add a return value
    - If you don't have a `return()` R will return the result of the last line

```{r}
#| output: false
busiest_routes = function () {
    pairs = group_by(data, Origin, Dest) %>% summarize(Passengers=sum(Passengers),
        distance_km=first(Distance) * 1.609)
    arrange(pairs, -Passengers)

    pairs = mutate(pairs,
        airport1 = if_else(Origin < Dest, Origin, Dest),
        airport2 = if_else(Origin < Dest, Dest, Origin)
        )
    pairs = group_by(pairs, airport1, airport2) %>% summarize(
        Passengers=sum(Passengers),
        distance_km=first(distance_km)
    )
    
    return(arrange(pairs, -Passengers))
}
```

# Running our function

- Running the code on the previous slide defines the function, but does not run it
- We have to _call_ the function to run it

```{r}
busiest_routes()
```

# Adding arguments

- We want to make the `Origin` and `Dest` column names arguments to the function
- It's also standard to make the data itself and anything else you refer to inside the function arguments

# Adding arguments

```{r}
#| output: false
busiest_routes = function (dataframe, origcol, destcol) {
    pairs = group_by(dataframe, origcol, destcol) %>% summarize(Passengers=sum(Passengers),
        distance_km=first(Distance) * 1.609)
    arrange(pairs, -Passengers)

    pairs = mutate(pairs,
        airport1 = if_else(origcol < destcol, origcol, destcol),
        airport2 = if_else(origcol < destcol, destcol, origcol)
        )
    pairs = group_by(pairs, airport1, airport2) %>% summarize(
        Passengers=sum(Passengers),
        distance_km=first(distance_km)
    )
    
    return(arrange(pairs, -Passengers))
}
```

# Adding arguments

```{r}
#| error: true
busiest_routes(data, Origin, Dest)
```

# What happened?

- R tried to group by the columns `origcol` and `destcol`
- These columns don't exist
- We need to tell R that `origcol` and `destcol` aren't actually column names, but variables that contain column names
- Any place we refer to a variable containing a column name inside a function, use the "embracing operator" `{{ variable }}` to tell R to use the value contained by `variable` as the column name
    - This is only true when the variable name would not usually be in quotes (`" or '`)

# Adding arguments

```{r}
#| output: false
busiest_routes = function (dataframe, origcol, destcol) {
    pairs = group_by(dataframe, {{ origcol }}, {{ destcol }}) %>% summarize(Passengers=sum(Passengers),
        distance_km=first(Distance) * 1.609)
    arrange(pairs, -Passengers)

    pairs = mutate(pairs,
        airport1 = if_else({{ origcol }} < {{ destcol }}, {{ origcol }}, {{ destcol }}),
        airport2 = if_else({{ origcol }} < {{ destcol }}, {{ destcol }}, {{ origcol }})
        )
    pairs = group_by(pairs, airport1, airport2) %>% summarize(
        Passengers=sum(Passengers),
        distance_km=first(distance_km)
    )
    
    return(arrange(pairs, -Passengers))
}
```

# Adding arguments

```{r}
busiest_routes(data, Origin, Dest)
```

# Using our function for cities

```{r}
busiest_routes(data, OriginCity, DestCity)
```

# Putting it all together

```{r}
#| output: false

busiest_routes = function (dataframe, origcol, destcol) {
    pairs = group_by(dataframe, {{ origcol }}, {{ destcol }}) %>% summarize(Passengers=sum(Passengers),
        distance_km=first(Distance) * 1.609)
    arrange(pairs, -Passengers)

    pairs = mutate(pairs,
        airport1 = if_else({{ origcol }} < {{ destcol }}, {{ origcol }}, {{ destcol }}),
        airport2 = if_else({{ origcol }} < {{ destcol }}, {{ destcol }}, {{ origcol }})
        )
    pairs = group_by(pairs, airport1, airport2) %>% summarize(
        Passengers=sum(Passengers),
        distance_km=first(distance_km)
    )
    
    return(arrange(pairs, -Passengers))
}

busiest_routes(data, Origin, Dest)
busiest_routes(data, OriginCity, DestCity)
```

# Exercise: do the same in `airline_analysis.R`

- Write a function to compute market shares
- Use that function to compute market shares by operating carrier and ticketing carrier
- Bonus: also compute market shares for individual airports instead of cities

# Result

```{r}
market_shares = function (dataframe, carrier, city) {
  mkt_shares = group_by(dataframe, {{ carrier }}, {{ city }}) %>%
    summarize(Passengers=sum(Passengers)) %>%
    group_by({{ city }}) %>%
    mutate(market_share=Passengers/sum(Passengers), total_passengers=sum(Passengers)) %>%
    ungroup()

  res = filter(mkt_shares, total_passengers > 1000) %>% arrange(-market_share)
  return(res)
}

```

# Result

```{r}
market_shares(data, OperatingCarrierName, OriginCity)
market_shares(data, TicketingCarrierName, OriginCity)
market_shares(data, TicketingCarrierName, Origin)
```


# Symbolic constants

- We still have the value `1.609` in our function to convert miles to kilometers
- It's a bad practice to have literal values like these in your code
    - In a large project, value will end up spread over many files
    - Difficult to change (maybe you want to increase accuracy later by changing it to `1.609344`)
    - Numbers used in many places can be prone to typos

# Symbolic constants

- We can define a _constant variable_ in our file that contains the value `1.609`, and refer to it when needed
- By convention, constants are all uppercase, and usually appear near the start of the file
- R does not enforce variables being constant, you could change them in other code, but you shouldn't

# Symbolic constants

- Near the top of your code file, add

```{r}
KILOMETERS_PER_MILE = 1.609
```

- Replace `1.609` in your function with `KILOMETERS_PER_MILE`

# Runtime checks: `stopifnot`

- A first step towards ensuring correctness is to add checks throughout your code
- In R, this is most easily done with the `stopifnot` function - this will error if whatever condition it checks is not true

# Runtime checks: `stopifnot`

- Let's make sure all of the passenger numbers are positive and non-NA
- Add these lines to the start of your `busiest_routes` function:
```{r}
#| eval: false
stopifnot(all(dataframe$Passengers >= 1))
stopifnot(all(!is.na(dataframe$Passengers)))
```

```{r}
#| include: false
# redefine to add stopifnot
busiest_routes = function (dataframe, origcol, destcol) {
  stopifnot(all(dataframe$Passengers >= 1))
  stopifnot(all(!is.na(dataframe$Passengers)))
  pairs = group_by(dataframe, {{ origcol }}, {{ destcol }}) %>% summarize(Passengers=sum(Passengers), distance_km=first(Distance) * KILOMETERS_PER_MILE)
  arrange(pairs, -Passengers)
  
  # we see that LAX-JFK (Los Angeles to New York Kennedy) is represented separately
  # from JFK-LAX. We'd like to combine these two. Create airport1 and airport2 fields
  # with the first and second airport in alphabetical order.
  pairs = mutate(pairs, airport1 = if_else({{ origcol }} < {{ destcol }}, {{ origcol }}, {{ destcol }}), airport2 = if_else({{ origcol }} < {{ destcol }}, {{ destcol }}, {{ origcol }}))
  pairs = group_by(pairs, airport1, airport2) %>% summarize(Passengers=sum(Passengers), distance_km=first(distance_km))
  return(arrange(pairs, -Passengers))
}
```

# Confirming that errors occur

```{r}
#| error: true
busiest_routes(mutate(data, Passengers=-1), Origin, Dest)
```

# Using functions in multiple files

- Often we may want to use functions in multiple files
- For instance, we use the same data loading code in both `air_analysis.R` and `airline_analysis.R`
- The simplest way is to put the functions in another file, and use `source()` to load them
- `source()` will run the code in a file and make all of the functions and variables available
- Just as if that code was pasted into each source file

# Creating a data cleaning function

- Create a function that takes arguments for the file names of the dataset and the two tables that are joined to it,
- performs the data cleaning/processing,
- and returns the final dataset
- Save this function to a new file in the same directory as the others

# Using your data cleaning function

- Delete the data loading code in each of the analysis files
- Use the 🧹 button in RStudio to clear the data from memory
- Add and run a line `source("filename.R")` to load your new R file
- Call your new function like this

```{r}
#| eval: false
data = load_data("data/air_sample.csv", "data/L_CITY_MARKET_ID.csv", "data/L_CARRIERS.csv")
```

# The finished function

Yours might not look exactly the same, and that's okay

```{r}
#| output: false
load_data = function(data_file, city_file, carrier_file) {
    data = read_csv(data_file)

    market_ids = read_csv(city_file)
    data = left_join(data, rename(market_ids, OriginCity="Description"),
        by=c(OriginCityMarketID="Code"))
    data = left_join(data, rename(market_ids, DestCity="Description"),
        by=c(DestCityMarketID="Code"))

    carriers = read_csv(carrier_file)
    data = left_join(data, rename(carriers, OperatingCarrierName="Description"),
        by=c(OpCarrier="Code"))
    data = left_join(data, rename(carriers, TicketingCarrierName="Description"),
        by=c(TkCarrier="Code"))

    return(data)
}
```

```{r}
#| include: false
# run this just to make sure function above works
rm(data)

data = load_data("../data/air_sample.csv", "../data/L_CITY_MARKET_ID.csv", "../data/L_CARRIERS.csv")
```

# All data manipulation should be in code

- I _never_ modify original data files, and rarely save new ones
- I always write code or functions to make all the changes/cleaning I need
- I run that function to load the data each time
- This way I can always see what has been done to data

# Creating R packages

- You use packages all the time when you use R
    - Any time you run `library(...)` you're loading a package
- You can also create packages yourself

# Advantages of creating a package

- Easy to share code, even between projects (no copying .R files to load with `source()`)
- Packages facilitate automated testing

# Creating an R package

- Several packages are available that make creating your own packages _much_ easier
    - `devtools` for overall workflow management
    - `roxygen2` for documentation generation and 
    - `testthat` for automated testing
- Install these now
```{r}
#| eval: false
install.packages(c("devtools", "roxygen2", "testthat"))
```

# What is an R package?

- A collection of functions, data, and documentation that can be easily installed in R and available to all R scripts

# Structure of an R package

- R packages are created in _source_ form
- This just means that they consist of the original R code
- When you install a package, it is converted to _binary_ form
    - Only readable by the computer, and ready to be run by R

# Structure of an R package

An R source package is just a directory with a specific structure

- `packagename`
    - `packagename.Rproj`
    - `.Rbuildignore`
    - `DESCRIPTION`
    - `NAMESPACE`
    - `R`
        - Your R files

# Creating your R package

- Load `devtools` and use `create_package(...)` to create the directory structure for your package
- You should put the package in a directory that is separate from your main project directory
- `create_package(...)` will create the directory if it doesn't already exist
- Call your package `airlinestats`
- A new RStudio window will open with your package

# Creating your R package

```{r}
#| eval: false
library(devtools)
create_package("~/git/airlinestats")
```

- A new RStudio window will open with the code for your new package

# Adding R code to your package

- Your R code goes in files in the `R/` directory
- Any file in this directory ending in `.R` will be part of your package

# Sidebar: namespaces

- Multiple packages may define functions with the same name
- To keep this straight, R uses _namespaces_
- Each package gets a namespace, and functions are defined within that namespace
- When you use `library(package)`, functions from that package are loaded into the global (main) namespace
- `package::function` refers to a function within a specific package's namespace

# Differences between R scripts and package code

- Packages should never use `library()` as this could affect code from other packages
- Almost everything in a package should be inside a function
    - Code that is not in functions will be run when the package is compiled or installed, not when it is used
    - Main exception is constants

# Adding R code to your `airlinestats` package

- Create two `.R` files, `load.R` and `analysis.R`
- Put your `load_data` function in `load.R`
- Put the functions to calculate market shares and route popularity in `analysis.R`
- Put the symbolic constant `KILOMETERS_PER_MILE` in `analysis.R`
- _Don't_ include `library(tidyverse)` anywhere

# Loading your package

- During development, you can load your package with `load_all()`
- This will reload the latest version of all of your R code
- `load_all` is from `devtools`, so you have to load that first

```{r}
#| eval: false
library(devtools)
load_all()
```

# Using your package

- Functions from your package are now available
- Try running the `load_data` function

```{r}
#| eval: false
data = load_data(
    "../odum-modular-design/data/air_sample.csv",
    "../odum-modular-design/data/L_CITY_MARKET_ID.csv",
    "../odum-modular-design/data/L_CARRIERS.csv"
)
```

. . . 

<pre>Error in read_csv(data_file) : could not find function "read_csv"</pre>

# Referring to functions from other libraries

- Normally, `read_csv` and other functions become available when you run `library(tidyverse)`
- But we don't do this in libraries
- There are two main ways to refer to functions from other libraries
    - Fully-qualified notation: `readr::read_csv`
    - `@importFrom` comments

# Fully-qualified notation

- When using fully qualified notation, we put the name of the library followed by `::` in front of the function name
- So, `read_csv` is replaced with `readr::read_csv`
- `left_join` and `rename` are prefixed with `dplyr::`

# Fully qualified notation

- We tell R that we want to use functions from the libraries `readr` and `dplyr` by running `use_package()`
- Do this, and run `load_all` again, and confirm that `load_data` now works

```{r}
#| eval: false
use_package("readr")
use_package("dplyr")
```

# Using `tidyverse` functionality

- `tidyverse` is a collection of many packages
- Using `tidyverse` directly in libraries is discourages
- Most data manipulation functions are from `dplyr`, data reading functions from `readr`

# `@importFrom`

- If you don't want to type out `dplyr::` over and over, you can use `@importFrom` to bring a function into your package's namespace
- The `devtools` `use_import_from` function automates this task

# `@importFrom`

- Let's import `summarize`, `group_by`, `mutate`, `first`, `if_else`, `arrange`, `ungroup`, `filter` and `summarize` from `dplyr` to make our analysis code work
- We alse need to import `%>%` from `magrittr`

```{r}
#| eval: false
use_import_from("dplyr", c("summarize", "group_by", "mutate", "first", "if_else", "filter", "ungroup", "arrange", "summarize"))
use_import_from("magrittr", "%>%")
```

- Accept the prompt to generate package documentation

# What happened?

- `devtools` created a new `airlinestats-package.R` that contains `@importFrom` statements to bring these functions into the package namespace
- The file consists of comments, but the `roxygen2` documentation system will convert these comments into import statements in the file `NAMESPACE` when the documentation is built

# Confirm it worked

- Re-run `load_all`
- Run the two analyses again:

```{r}
#| eval: false
market_shares(data, OperatingCarrierName, OriginCity)
busiest_routes(data, OriginCity, DestCity)
```

# Checking the package

- Run the `check()` function from `devtools`

# Checking the package

```{r}
#| eval: false
check()
```

. . .

<pre>
❯ checking DESCRIPTION meta-information ... WARNING
  Non-standard license specification:
    `use_mit_license()`, `use_gpl3_license()` or friends to pick a
    license
  Standardizable: FALSE

❯ checking R code for possible problems ... NOTE
  busiest_routes: no visible binding for global variable ‘Passengers’
  busiest_routes: no visible binding for global variable ‘Distance’
  busiest_routes: no visible binding for global variable ‘airport1’
  busiest_routes: no visible binding for global variable ‘airport2’
  busiest_routes: no visible binding for global variable ‘distance_km’
  market_shares: no visible binding for global variable ‘Passengers’
  market_shares: no visible binding for global variable ‘market_share’
  Undefined global functions or variables:
    Distance Passengers airport1 airport2 distance_km market_share

0 errors ✔ | 1 warning ✖ | 1 note ✖
</pre>

# Addressing the issues

- `Non-standard license specification`
    - R is suggesting you fill out the `DESCRIPTION` file, which includes choosing a license. We'll skip this for now.
- `No visible branding` and `undefined global`
    - These are the columns of our data; R doesn't know this when checking the code, though, and warns that they may be undefined

# Installing the package

- Just run `install()` to install the package on your machine
- Now it will be available in any R script
- You may be prompted to update packages when you do this

# Bringing it full circle

- Return to the original script files
- Replace the code in `air_analysis.R` and confirm you get the same results

```{r}
#| eval: false
library(airlinestats)
data = load_data("data/air_sample.csv", "data/L_CITY_MARKET_ID.csv", "data/L_CARRIERS.csv")
busiest_routes(data, Origin, Dest)
busiest_routes(data, OriginCity, DestCity)
```

. . . 

<pre>
Error in load_data("data/air_sample.csv", "data/L_CARRIERS.csv", "data/L_CITY_MARKET_ID.csv") : 
  could not find function "load_data"
</pre>

# Fixing the error

- Only functions you explicitly _export_ from a package will be available outside the package
- This allows you to have functions used only internally
- We mark a function to be exported by adding a comment above the function definition
```{r}
#' @export
```
- Mark all three functions exported

# Reinstalling the package

- We need to run three functions to reinstall
    - `document()` to parse the special comments we added
    - `check()` to help ensure we didn't introduce additional errors
    - `install()` to reinstall the package

# A new warning

<pre>
❯ checking for missing documentation entries ... WARNING
  Undocumented code objects:
    ‘busiest_routes’ ‘load_data’ ‘market_shares’
  All user-level objects in a package should have documentation entries.
  See chapter ‘Writing R documentation files’ in the ‘Writing R
  Extensions’ manual.
</pre>

- R suggests that all exported functions in a package have documentation
- We'll ignore this for now

# Finishing up

- Re-run the code using your package; it should work now
- You may have to restart R

# Automated testing

- Automated testing formalizes testing and makes it repeatable
- Every test is run automatically when you run `check()`
- If you break something later, you are more likely to find out about it if your code has automated tests

# The structure of testing

- Automated tests consist of code and _expectations_
    - Expectations are called _assertions_ in other languages
- A test exercises one of the functions in your package, and specifies expectations about its output
- Good tests are small, with one piece of code tested and a handful of expections
- This may require splitting your package code up into multiple functions
- It's a good idea to test how your package acts when used incorrectly as well—for instance, when fed bad data

# Automated testing in R

- In R, automated testing is provided by the `testthat` package
- We can add testing to our package by running

```{r}
#| eval: false
use_testthat(3)
```

# Automated testing in R

- We now have a `tests/` directory, with a `testthat` directory inside it
- Any file starting with `test` and ending with `.R` in this directory will be run as part of automated testing

# Writing tests

- Create an R file to hold tests for the analysis functions
- It's good practice to match the names of the test files to the names of the files that define the code being tested, so create `tests/testthat/test-analysis.R`

# Writing tests

- We'll write two tests for the analysis code
    1. We'll test that market shares of all airline at an airport sum to 1
    2. We'll test that passing in a dataset without a `Passengers` column leads to an error

# Writing tests

- Often, we need data for our tests
- Here, we will create a small dataset within our test file directly
- It is also possible to distribute a CSV with your package with test data
- I like to put my code to generate test data in a function, and re-create the data for each test, so that there is no chance of one test modifying the data used by another
    - Unless the data take so long to create or load that this is impractical

# Test data

```{r}
#| eval: false
get_data = function () {
  data = dplyr::tibble(
    Origin=rep(c("SFO", "ORD", "DCA", "PHX", "BOS", "RDU"), 20),
    Destination=rep(c("DFW", "MCI", "MIA", "ABQ", "IAH", "SJU"), 20),
    Passengers=rep(1:6, 20),
    Carrier=rep(c("United", "American", "Delta", "Southwest"), 30)
  )
  return(data)
}
```

# Write the test for market shares

```{r}
#| eval: false
test_that("Market shares sum to 1", {
  data = get_data()
  shares = market_shares(data, Carrier, Origin)
  total_shares = group_by(shares, Origin) %>% summarize(overall=sum(market_share))
  # six airports, so six shares
  expect_equal(total_shares$overall, rep(1, 6))
})
```

# Write the test for bad input

```{r}
#| eval: false
test_that("Error when passengers column missing", {
  data = get_data() %>% dplyr::select(-Passengers)
  expect_error(busiest_routes(data, Origin, Destination))
})
```

# Run tests

- Click "run tests" in the upper-right corner in RStudio
    - This will run the tests in the current test file
- Alternately, run
```{r}
#| eval: false
test_file("tests/testthat/test-analysis.R")
```

# Fix test errors

- We got a test failure for the market shares test
- This is because we filtered out small airlines in the function
- Let's remove the filtering from the function
    - It's good to break your functions into smaller pieces for easier testing

# Exercise

- Check and reinstall your package
- Notice that your tests were run
- Confirm your scripts still work

# When...

## to create a package

- When you are building reusable functions for a large project, or several projects
- When you want to employ automated testing

## not to create a package

- Simple analyses and one-off projects
- Projects with few interdependencies and where `stopifnot` testing is sufficient

# Environment management and `renv`

- R projects rely heavily on packages
- Changes to package versions may break code or change results
- Reproducibility requires keeping track of used packages and versions

# `renv`

- `renv` is a package that automates this
- Once installed, set up a new project with `renv::init()` and snapshots current packages in use in a _lockfile_
- `renv::snapshot()` updates currently used packages
- `renv::restore()` will reinstall dependencies on a new machine or years in the future

# `renv`

- You should use `renv` with your scripts, not in packages
- `renv` will record the versions of packages you use

# Exercise: set up `renv` in your project

- Install `renv` if needed
- Run `renv::init()`
- Inspect `renv.lock`

# `renv` and package development

- `install()` installs your package system-wide
- `renv` uses its own per-project library of packages
- To re-install your custom package, you need to run `renv::install("path/to/package/directory")` from within your `renv` environment (i.e. your project with scripts, not your package)
- Otherwise you will always have the version you had when you first ran `renv::init`
- Since we haven't discussed versioning or deploying packages to CRAN, `renv` can't track custom packages

# Additional topics

- Version control
- Continuous integration
- Package versioning
- Publishing to CRAN

# Additional resources

- R package book: [r-pkgs.org](https://r-pkgs.org/)

## Contact

Matt Bhagat-Conway<br/>
[mwbc@unc.edu](mailto:mwbc@unc.edu)